---
title: "org.Hs.eg.db3: parquet-based AnnotationDbi for Hs"
shorttitle: "outlier dytection methods"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{org.Hs.eg.db3: parquet-based AnnotationDbi for Hs}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

There are two key characteristics of the org.Xx packages.

- environments that manage mappings between identifiers;
- database tables that respond to AnnotationDbi::select and possibly other
SQLite-oriented processes.

We are moving away from SQLite and the Annotation pipeline.  This
package is intended to be more maintainable.

# Environment example

In the AnnotationDbi approach, we have various methods for
extracting mappings between identifiers.  We preserve those
methods here.

```{r lkenv1, message=FALSE}
library(org.Hs.eg.db3)
org.Hs.egGO
class(org.Hs.egGO)
str(get("26538", org.Hs.egGO)[1:3])
toTable(org.Hs.egGO) |> head(10)
```

There is also `AnnotationDbi::select`, which has served very well.

# Parquet example

However, the most basic information from NCBI can also
be presented directly.

```{r lkparq1}
path = system.file("extdata", "Hs.gene2go.parquet", package="org.Hs.eg.db3")
ggtab = arrow::read_parquet(path)
ggtab |> head()
```

We use parquet for the serialization because it can foster cloud-oriented
methods for annotation servicing.  With this approach, tidyverse methods
are immediately applicable; this is not so for the AnnotationDbi approach.

In this approach, we do not need separate environments for GO and GO2EG.
Similarly, separate environments for mapping REFSEQ, REFSEQ2EG, SYMBOL,
SYMBOL2EG are not required.  We provide them to prevent legacy code from
breaking, but we will not provide them indefinitely.

# Creating the parquet

NCBI frequently updates mapping tables in text format.  The key URL
as of March 2026 is
```
https://ftp.ncbi.nlm.nih.gov/gene/DATA/
```
We have
```
Parent Directory                                     -   
ARCHIVE/                        2020-01-02 15:38    -   
ASN_BINARY/                     2026-03-12 20:32    -   
GENE_INFO/                      2026-03-12 20:32    -   
expression/                     2017-03-06 17:55    -   
special_requests/               2025-02-28 11:35    -   
README                          2025-01-02 08:35   58K  
README_ensembl                  2025-12-03 12:00   69K  
gene2accession.gz               2026-03-12 20:18  4.0G  
gene2ensembl.gz                 2026-03-12 20:18  277M  
gene2go.gz                      2026-03-12 20:20  1.2G  
gene2pubmed.gz                  2026-03-12 20:20  245M  
gene2refseq.gz                  2026-03-12 20:23  2.0G  
...
```

The following code operates on the `gene2ensembl.gz` (after downloading) to
create a data resource *for all model orgnisms*:
```
library(DBI)
library(duckdb)

con = dbConnect(duckdb())

dbExecute(con, "
  CREATE TABLE gene2ensembl AS
  SELECT *
  FROM read_csv_auto(
    'gene2ensembl.gz',
    delim = '\t',
    header = true,
    compression = 'gzip',
    ignore_errors=TRUE
  )
")

dbExecute(con, "COPY (FROM gene2ensembl) TO 'gene2ensembl.parquet' 
       (FORMAT parquet, COMPRESSION zstd, COMPRESSION_LEVEL 15)")
```

# Creating the extract for human

```
library(arrow)
ens = read_parquet("gene2ensembl.parquet")
library(dplyr)
ens |> filter(`#tax_id`==9606L) |> 
  write_parquet(sink="Hs.gene2ensembl.parquet", 
  compression="zstd", compression_level=10L)
```

This is then placed in the extdata folder for an org.Hs.eg.db-like package.

# Creating the environments to support legacy usage

For the task of the creation of a map that has the form of a
named list of character vectors, we have

```
simpleEnvBuilder = function(keys, values, keycolname, valcolname) {
      force(keys)  # this is crucial to avoid delayed evaluation until active binding is triggered
      force(values)
      force(keycolname)
      force(valcolname)
     function(pkgname, target_env_name = "org.Hs.egENSEMBL") {
  stopifnot(is.character(keys))
  stopifnot(is.list(values))
  CURRENT = target_env_name
  CURE = paste0(target_env_name, "_env")
  ns = getNamespace(pkgname)
  assign(CURE, new.env(hash=TRUE), envir=ns)
  nn = lapply(seq_len(length(keys)), function(i) assign(keys[i], values[[i]], envir=get(CURE, envir=ns)))
  f = function() {
    tmp = get(CURE, envir=ns)
    class(tmp) = c("hsParqEnv", "environment")
    attr(tmp, "keycolname") = keycolname
    attr(tmp, "valcolname") = valcolname
    tmp
    }
  rm(list=CURRENT, envir=ns)
  makeActiveBinding(CURRENT, f, ns)
 }
}
```

This produces environments that function
like those of the AnnotationDbi packages.

# Omitted environments

The full set of environments offered by the org.Hs.eg.db in
Bioc 3.22 is
```
 [7] "org.Hs.egACCNUM"          "org.Hs.egACCNUM2EG"       "org.Hs.egALIAS2EG"       
[10] "org.Hs.egCHR"             "org.Hs.egCHRLENGTHS"      "org.Hs.egCHRLOC"         
[13] "org.Hs.egCHRLOCEND"       "org.Hs.egENSEMBL"         "org.Hs.egENSEMBL2EG"     
[16] "org.Hs.egENSEMBLPROT"     "org.Hs.egENSEMBLPROT2EG"  "org.Hs.egENSEMBLTRANS"   
[19] "org.Hs.egENSEMBLTRANS2EG" "org.Hs.egENZYME"          "org.Hs.egENZYME2EG"      
[22] "org.Hs.egGENENAME"        "org.Hs.egGENETYPE"        "org.Hs.egGO"             
[25] "org.Hs.egGO2ALLEGS"       "org.Hs.egGO2EG"           "org.Hs.egMAP"            
[28] "org.Hs.egMAP2EG"          "org.Hs.egMAPCOUNTS"       "org.Hs.egOMIM"           
[31] "org.Hs.egOMIM2EG"         "org.Hs.egORGANISM"        "org.Hs.egPATH"           
[34] "org.Hs.egPATH2EG"         "org.Hs.egPFAM"            "org.Hs.egPMID"           
[37] "org.Hs.egPMID2EG"         "org.Hs.egPROSITE"         "org.Hs.egREFSEQ"         
[40] "org.Hs.egREFSEQ2EG"       "org.Hs.egSYMBOL"          "org.Hs.egSYMBOL2EG"      
[43] "org.Hs.egUCSCKG"          "org.Hs.egUNIPROT"        
```

- the ACCNUM environments refer to Genbank and are not updated by NCBI; they are omitted in 3.23
- the `CHR*` environments should not be retrieved from this package and are omitted
- the `ENZYME and PATH environments refer to old versions of KEGG and are omitted

# Session information

```{r dosess}
sessionInfo()
```
